# Introduction to Clinical Prediction Models {#intro}

Here is a review of Clinical Prediction Models (on the clinical aspect)

## Prediction/prognosis in clinical medicine

In daily clinical practice, one of the main tasks of clinicians is to investigate and develop three main areas of knowledge related to the patient's illness: diagnosis (whether a particular illness is present), aetiology (what is the cause of the present illness) and prognosis (what is the likely future course of the present illness) [@Miettinen2011]. Based on this, clinicians can suggest a suitable intervention or management plan for the patient under investigation. Therefore, making prognostic assessments is a natural and familiar task that clinicians have to do every day.

As prediction is a fundamental component of medical decision-making, it has an important role in clinical practice. However, it is also a generally difficult task because unlike diagnosis or etiognosis, prognosis requires extrapolation to the future based on the present knowledge about a patient's disease status and characteristics. As the future is uncertain, predictions need to rely on many assumptions and making a good prediction is an extremely difficult task.

In general, the prediction making process may come with pitfalls including its subjectivity, and proneness to errors and inconsistencies [@Meehl1954, @Dawes1989, @Grove2000, @Sox2013]. Common pitfalls when estimating probability using only personal experience are:

* Focusing on the presence or absence of predictors but ignoring the prior probability of the outcome.

* Basing predictions on the presence of predictors with low predictive ability or a set of predictors which do not independently affect outcome.

* Personal experience, especially for clinicians in their early careers, is typically a small and unrepresentative sample of the overall population.

* Reliance on a false belief in the relationship between candidate predictors and outcome, as it is difficult to differentiate invalid relationships from valid ones theoretically.

* A tendency to be over-confident and overstate subjective certainty.

As a result, deriving precise and unbiased predictions from personal experience alone is beyond the cognitive ability of almost everyone [@Sox2013]. Therefore, additional knowledge, especially that deriving from an objective source like empirical research, is needed. Such knowledge includes published reports describing the incidence of outcomes amongst patients having a common set of clinical features and clinical prediction models. Compared to personal experience, they provide more objective and more comprehensive information, especially for rare outcomes. However, their utility and validity can vary widely depending on the quality of the clinical studies and the similarities between the research population and the population of interest. Amongst these objective sources of knowledge, clinical prediction models are recognized as powerful tools to derive prediction. Besides their ability to process and produce complex information that may go beyond human mental ability, clinical prediction models provide consistent estimates and have been shown to outperform personal judgment in many situations [@Dawes1989, @Grove2000]. Caveats in applying clinical prediction model are related to their potential to over-fit the data when they are not developed or tested properly and their potential to miss predictors which are relevant but difficult to evaluate in clinical research. 

Based on these rationales, a better model for estimating the risk of the occurrence of a certain outcome in clinical practice requires both published evidence and personal experience. In addition, further adjustments to risk estimation may be required due to differences between the population of interest and the reference population, or due to the presence of extraordinary clinical features in the patient that the physician has never seen and which have not been reported anywhere [@Sox2013].

## What are prognostic models?

A prognostic model is a formal combination rule of multiple predictors from which a subject’s absolute risk of the occurrence of a disease event of interest can be calculated [@Steyerberg2013]. Prognostic modeling is an important part of prognostic research which aims to understand and improve future outcomes in subjects with a given health state [@Hemingway2013]. Developing a prognostic model is the third step of a 4-step paradigm for prognostic research which includes (1) investigating the variation of outcomes of a health condition in the context of current care (fundamental prognostic research); (2) identifying risk factors which are related to outcomes (prognostic factor research); (3) developing, validating and assessing the impact of prediction models that predict an individual’s risk of a future event (prognostic model research); (4) using prognostic information to help individualize treatment decisions for a subject or group of subjects that share similar characteristics (stratified medicine research) [@Hemingway2013].

A typical prediction model has three main ingredients: an outcome, candidate predictors, and a statistical model. An outcome can be a hard endpoint such as death or the presence of clinical complication, or a surrogate endpoint such as a biomarker of severity, or a composite endpoint. Candidate predictors typically include demographic variables such as age and gender as well as clinical symptoms or biomarkers which are relevant to outcome prediction based on clinical knowledge. As prognosis is aimed to the future, predictors have to be collected at a starting point or baseline which is before the outcome occurs and the length of the lag period between the starting point and the outcome occurrence affects largely the usefulness of the derived prediction model. The relationship between outcome and predictors is modeled using a statistical model and the choice of this model depends on the type of outcome. As clinical outcomes are usually binary or survival data, the most common statistical models of choice are logistic regression and the Cox proportional hazards model, but many other statistical models have been suggested in the literature [@Steyerberg2010,@Hastie2009]. As a general rule, these three main components have to be pre-specified before fitting any model in order to avoid over-fitting and therefore to preserve the validity of the derived model.

Prediction models are different from decision rules. The inputs to a prediction model are values of prognostic factors at a pre-defined time point and the output is an estimated risk of a specific outcome. Even though one may categorize estimated risks and assign suggested actions to each risk category, a typical prediction model only provides a prediction of a risk as it is intended to assist clinicians without suggesting to them what to do [@Reilly2006a]. In contrast, a decision rule is designed to directly affect clinical decisions by physicians. As accurate predictions do not always improve clinical decisions, a promising prediction model has to demonstrate its positive impact on physicians’ decisions and patients’ outcomes through different levels of impact analysis in order to be successfully translated into a useful clinical decision rule [@Reilly2006a].

As described in the previous section, the major advantage of prediction models in clinical practice is their ability to provide objective, reproducible and reliable estimation of outcome occurrence. Due to their transparency, prediction models can also enhance communication between physicians and patients [@Steyerberg2013]. Moreover, in clinical research prediction models can be used in the design stage to target a population of interest, and in the analysis stage to perform stratified analysis according to predicted risk groups or to improve the power of statistical analyses [@Steyerberg2013].

#### Baseline versus dynamic prediction models

Traditionally, a prediction models relies on data collected at a single time point (at the time of presentation, admission, diagnosis or initiation of an intervention) to predict outcomes in the future. Even though many of these traditional prediction models are useful in clinical practice, they have several shortcomings:

* Initial predictions tend to become less relevant as the disease progresses [@Rue2001]. A possible explanation is that a prediction model based on baseline information only cannot capture changes in the patient's clinical profile according to their response to treatment or natural physiologic variation reflecting the course of the disease, which may be strong predictors of the outcome [@Rue2001]. Furthermore, baseline models might also miss complications, which may strongly affect outcome, while being present at baseline but require time to become clinically apparent [@Lemeshow1988]. As a result, baseline models might not be transferable to later time points and therefore, they may not be used for individual management decisions at those time points [@Lemeshow1994, @Wagner1994].

*	Longitudinal information during the patient’s disease course is nowadays frequently collected in clinical practice, especially with the introduction of electronic health record. Baseline models are inefficient in the sense that they ignore all this subsequent after baseline.

For these reasons, dynamic prediction models, which predict the future course of the disease at follow-up time points based on the accruing longitudinal information, are required to allow updating a patients' prognosis over time [@VanHouwelingen2012]. By using all available data, such models may provide much more accurate predictions compared to baseline models in many settings [@Lemeshow1988, @Christensen1993, @Hughes1992, @Rue2001, @Karp2004]. Dynamic prediction may also be appealing for clinicians as it mimics the iteration of obtaining information and updating prognosis based on the new information, a task that physicians routinely do every day in clinical practice.  

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(CPM).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015]. ddd aaa
